---
title: "The Burger Standard: What a Hamburger Says About Cheap and Dear Currencies"
author: "Siyu Xin"
date: 2026-10-05
categories:
  - markets
  - FX
  - data visualization
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    df-print: kable
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(readr)      # read the data (direct from a URL)
library(dplyr)      # data wrangling
library(tidyr)      # reshaping
library(stringr)    # labels
library(ggplot2)    # figures
library(scales)     # axis formatting (%, $)
library(knitr)      # kable tables
library(here)       # project-relative paths

set.seed(6400)
options(scipen = 999)
dir.create(here("blog", "posts", "post4", "data"), showWarnings = FALSE, recursive = TRUE)
```

# The question

Exchange rates are supposed to tell you what a currency is worth, but on their own
they tell you almost nothing about whether a currency is *cheap* or *expensive*. A
dollar buys about 150 yen. Is that a lot or a little? The only way to answer is to
ask what the same thing costs in both places. In 1986 *The Economist* picked a
gloriously simple "same thing": a McDonald's Big Mac, sold in nearly every country
to roughly the same recipe. If a Big Mac costs far less in one country than in the
US once you convert at the market exchange rate, that country's currency looks
**undervalued**: your dollars go further than they "should."

This post uses 25 years of *Economist* Big Mac data to ask a question that is
sillier than it sounds and more serious than it looks: **which currencies does a
hamburger call cheap or dear, and does that verdict survive once you account for
how rich each country is?** The three figures build the answer, and the twist is in
the second one.

# The data and the one transformation that matters

The data come straight from *The Economist's* public [Big Mac
repository](https://github.com/TheEconomist/big-mac-data). Reading directly from its
raw CSV URL, so the whole analysis is reproducible from code. It covers 54
currencies, twice a year, back to 2000.

The entire index rests on one small transformation. For each country I take the
local Big Mac price, convert it to dollars at the market exchange rate, and compare
it to the US price:

$$\text{raw valuation} = \frac{\text{Big Mac price in USD}}{\text{US Big Mac price}} - 1$$

A value of −0.40 means a currency's burger is 40% cheaper than America's. The
currency is 40% "undervalued" against the dollar on this measure. That is the whole
idea: purchasing-power parity, reduced to a sandwich.

```{r}
#| label: load-data

# Reproducible fetch: pull once from the official repo, then cache a snapshot so
# every later render reads the same data offline. Set refresh <- TRUE to re-pull.
refresh   <- FALSE
data_url  <- "https://raw.githubusercontent.com/TheEconomist/big-mac-data/master/output-data/big-mac-full-index.csv"
snap_file <- here("blog", "posts", "post4", "data", "big-mac-full-index.csv")

if (!refresh && file.exists(snap_file)) {
  bmac_raw <- read_csv(snap_file, show_col_types = FALSE)
} else {
  bmac_raw <- read_csv(data_url, show_col_types = FALSE)
  write_csv(bmac_raw, snap_file)
}

# The US price for each date (the yardstick), then compute the raw valuation
# ourselves — it reproduces the Economist's USD_raw column exactly.
us_price <- bmac_raw |>
  filter(iso_a3 == "USA") |>
  select(date, us_price = dollar_price)

bmac <- bmac_raw |>
  left_join(us_price, by = "date") |>
  mutate(
    dollar_price = local_price / dollar_ex,       # burger price in USD
    val_raw      = dollar_price / us_price - 1,    # our transformation (= USD_raw)
    val_adj      = USD_adjusted                    # GDP-adjusted version (see Fig 2)
  )

latest     <- max(bmac$date)
latest_lab <- format(latest, "%B %Y")
```

```{r}
#| label: prose-stats
#| include: false

snap <- bmac |> filter(date == latest)
us_now     <- round(unique(snap$us_price), 2)
n_cur      <- n_distinct(snap$currency_code)

cheapest   <- snap |> filter(iso_a3 != "USA") |> slice_min(val_raw, n = 1)
dearest    <- snap |> slice_max(val_raw, n = 1)

bs_fit     <- lm(val_raw ~ GDP_bigmac, data = snap)
bs_slope   <- coef(bs_fit)[["GDP_bigmac"]] * 10000 * 100   # %pts per $10k GDP
bs_r       <- cor(snap$GDP_bigmac, snap$val_raw, use = "complete.obs")

jpy_now    <- bmac |> filter(currency_code == "JPY", date == latest)
jpy_raw_now<- jpy_now$val_raw * 100
jpy_adj_now<- jpy_now$val_adj * 100
jpy_2000   <- bmac |> filter(currency_code == "JPY") |> slice_min(date, n = 1)
jpy_raw_00 <- jpy_2000$val_raw * 100
```

# What the burger says

As of **`r latest_lab`**, an American Big Mac costs about **$`r us_now`**, and the
index covers **`r n_cur`** currencies. Figure 1 lines them all up from cheapest to
dearest against the dollar.

```{r}
#| label: fig-ranking
#| fig-cap: "Big Mac over/under-valuation against the US dollar, latest snapshot. Red bars = currencies a burger says are undervalued (your dollars go further); blue = overvalued."
#| fig-width: 7
#| fig-height: 8.5

snap |>
  filter(iso_a3 != "USA") |>
  mutate(name = reorder(name, val_raw)) |>
  ggplot(aes(val_raw, name, fill = val_raw > 0)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c(`FALSE` = "#c0392b", `TRUE` = "#2c7fb8"), guide = "none") +
  scale_x_continuous(labels = label_percent()) +
  labs(x = "Over / under-valuation vs. USD", y = NULL,
       title = paste0("What a burger says each currency is worth — ", latest_lab)) +
  theme_minimal(base_size = 9)
```

The spread is enormous: the **`r dearest$name`** franc looks roughly
**`r round(dearest$val_raw*100)`%** too dear, while the
**`r cheapest$name`** currency looks about **`r round(abs(cheapest$val_raw*100))`%**
too cheap. Taken at face value, that would mean half the emerging world is running
drastically "undervalued" currencies. But that face-value reading is a trap, and
Figure 2 is where it springs.

```{r}
#| label: fig-balassa
#| fig-cap: "The catch: richer countries have dearer burgers. Each point is a currency; the line is the average relationship. Points far below the line are cheap even for their income (genuinely undervalued) while points above are genuinely dear."
#| fig-width: 7.5
#| fig-height: 5

# only countries with GDP data can go on this chart
snap_bs <- snap |> filter(!is.na(GDP_bigmac), !is.na(val_raw))

lab_df <- snap_bs |>
  mutate(resid = resid(lm(val_raw ~ GDP_bigmac, data = snap_bs))) |>
  filter(currency_code %in% c("JPY", "CNY", "CHF", "NOK", "IDR", "INR") |
         abs(resid) >= quantile(abs(resid), 0.9, na.rm = TRUE))

ggplot(snap_bs, aes(GDP_bigmac, val_raw)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
  geom_smooth(method = "lm", se = FALSE, colour = "#2c3e50", linewidth = 0.8) +
  geom_point(colour = "#c0392b", alpha = 0.7, size = 2) +
  geom_text(data = lab_df, aes(label = name), size = 3, vjust = -0.8,
            check_overlap = TRUE) +
  scale_x_continuous(labels = label_dollar(scale = 1/1000, suffix = "k")) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = "GDP per capita", y = "Raw valuation vs. USD",
       title = "Cheap currency, or just a poor country?") +
  theme_minimal(base_size = 12)
```

The line slopes up: richer countries really do have pricier burgers, because wages,
rent and service costs (the non-traded ingredients) rise with income. Economists
call this the Balassa-Samuelson effect, and it means a poor country's "undervalued"
currency may be priced exactly right for how poor it is. So part of what Figure 1
calls cheapness is just income in disguise.

But only *part*. The relationship is loose: the correlation between income and
valuation is just **`r round(bs_r, 2)`**, a rise of about
**`r sprintf('%+.1f', bs_slope)`** points per extra $10k of GDP per capita. Income
nudges burger prices; it does not dictate them. The vertical distance from the line
is the piece income *can't* explain. The currency's genuine mispricing, which is
exactly what *The Economist's* "GDP-adjusted" index measures. In other words, the
burger survives the obvious objection: most of the cheapness is real.

Figure 3 puts one currency on the clock, the Japanese yen, the FX story of the
decade.

```{r}
#| label: fig-jpy
#| fig-cap: "The yen against the dollar on the Big Mac index, raw and GDP-adjusted, 2000–present. The yen went from overvalued at the start of the century to its most undervalued reading on record."
#| fig-width: 7.5
#| fig-height: 4.5

bmac |>
  filter(currency_code == "JPY") |>
  select(date, Raw = val_raw, `GDP-adjusted` = val_adj) |>
  pivot_longer(-date, names_to = "measure", values_to = "val") |>
  ggplot(aes(date, val, colour = measure)) +
  geom_hline(yintercept = 0, colour = "grey70", linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.4) +
  scale_colour_manual(values = c(Raw = "#c0392b", `GDP-adjusted` = "#2c3e50"),
                      name = NULL) +
  scale_y_continuous(labels = label_percent()) +
  labs(x = NULL, y = "Yen valuation vs. USD",
       title = "A burger tracks the yen's long slide") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

In 2000 the yen was actually **`r sprintf('%+.0f', jpy_raw_00)`%** *overvalued* on
the burger, a dollar bought a cheap American lunch and an expensive Tokyo one. By
**`r latest_lab`** the same measure reads about **`r round(jpy_raw_now)`%**
*undervalued*, the weakest in the 25-year series. The point that matters for an FX
view is that the GDP-adjusted line sits almost on top of the raw one, at roughly
**`r round(jpy_adj_now)`%**: Japan is a rich country, so Balassa-Samuelson says its
burger *should* be expensive, yet the yen is cheap anyway. This is not a poor-
country artifact. It is a genuinely, structurally weak currency, and a hamburger
spotted it without reading a single central-bank statement.

# Takeaway

The Big Mac index is a joke that keeps being right. Reduced to one sandwich,
purchasing-power parity produces a league table that mostly lines up with what FX
strategists believe: the Swiss franc is expensive, emerging-market currencies are
cheap, and the yen is historically, genuinely weak. The one discipline the burger
demands is the Balassa-Samuelson correction in Figure 2, without it you mistake
*poor* for *cheap*; with it you can separate the two.

The broader lesson is about transformation, not hamburgers. The raw data (a price
and an exchange rate) say nothing on their own. One division turns them into a
valuation; one scatter against income turns that valuation into an argument about
what's genuinely mispriced. That is the whole craft of this kind of analysis: the
numbers don't speak until you make them comparable.

---

*Data source: The Economist, Big Mac Index (`TheEconomist/big-mac-data` on GitHub,
MIT-licensed), read programmatically from its raw CSV and cached in `data/`.
Exchange rates and GDP-per-capita figures are those bundled in the dataset. Code to
reproduce the analysis is in the linked GitHub repository.*
