---
title: "What Is the Market Talking About? A Sentiment Read from Financial Headlines"
date: 2026-09-16
categories:
  - markets
  - web scraping
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    df-print: kable
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# --- packages -------------------------------------------------------------

library(tidytext); library(readr); library(here)
dir.create(here("data"), showWarnings = FALSE, recursive = TRUE)
write_csv(get_sentiments("loughran"), here("data", "loughran_lexicon.csv"))

library(rvest)      # read + parse the RSS (XML) feeds  <- the scraping tool
library(dplyr)      # data wrangling
library(tidyr)      # reshaping (pivot_wider)
library(stringr)    # text cleaning + keyword tagging
library(lubridate)  # parse publication dates
library(purrr)      # safe iteration over feeds
library(readr)      # write / read the data snapshot
library(tidytext)   # tokenising + Loughran-McDonald finance lexicon
library(ggplot2)    # figures
library(scales)     # percent() formatting
library(knitr)      # kable tables
library(here)       # project-relative paths (portable across machines)

# --- reproducibility settings --------------------------------------------
set.seed(6400)                 # nothing random here, but good habit
options(scipen = 999)

# --- project-relative folders (created if missing) -----------------------
dir.create(here("data", "raw"), showWarnings = FALSE, recursive = TRUE)
```

# The question

Every trading morning begins the same way: someone scans a wall of financial
headlines and forms a gut sense of *what the market cares about* and *whether the
mood is nervous or upbeat*. That instinct drives real decisions — what to flag to
clients, where the risk sits, what to read next. But a gut read is slow,
subjective, and impossible to compare from one day to the next.

This post asks a simple question: **can we turn the morning news scan into a
number?** From public financial-news headlines, can we measure (1) which themes
the market is talking about right now, and (2) whether coverage of each theme
leans positive or negative?

# The data and how it is collected

The raw material is headlines from public **RSS feeds** — the machine-readable
feeds that news sites publish specifically for programs to subscribe to. RSS is
plain XML, so the `rvest` package can read each feed and pull out every headline
and its timestamp. Because RSS exists *for* automated access, this sidesteps the
ethical and technical problems of scraping locked-down pages: there are no logins
or paywalls to bypass, and the code pauses one second between requests so it never
stresses a server.

```{r}
#| label: collect-headlines

# 1. The feeds we listen to (all public RSS 2.0 endpoints; no logins/paywalls).
#    If one is retired it is skipped automatically (see safe_parse below) —
#    check the printout and swap in any feed from the outlet's /rss page.
feeds <- tribble(
  ~source,              ~url,
  "CNBC Top News",      "https://www.cnbc.com/id/100003114/device/rss/rss.html",
  "CNBC Markets",       "https://www.cnbc.com/id/20910258/device/rss/rss.html",
  "CNBC Finance",       "https://www.cnbc.com/id/10000664/device/rss/rss.html",
  "Bloomberg Markets",  "https://feeds.bloomberg.com/markets/news.rss",
  "Bloomberg Business", "https://feeds.bloomberg.com/business/news.rss",
  "Bloomberg Tech",     "https://feeds.bloomberg.com/technology/news.rss",
  "MarketWatch Top",    "http://feeds.marketwatch.com/marketwatch/topstories/"
  # To add Investopedia: its site blocks bots and its native RSS is routed
  # through Dotdash, so open its Markets News page in a browser, copy the feed
  # URL, and uncomment the line below (leading comma keeps the tribble valid):
  # , "Investopedia News", "https://www.investopedia.com/PASTE_FEED_URL_HERE"
)

# 2. Politely fetch ONE feed and parse it into a tidy tibble.
#    RSS 2.0 structure is: channel > item > (title, pubDate, guid, ...).
#    NB: rvest's HTML parser lowercases tags, so we select "pubdate", not "pubDate".
#    We read the article URL from <guid> rather than <link>, because <link> is a
#    void element in HTML and read_html() drops its text.
parse_feed <- function(source, url) {
  Sys.sleep(1)                                   # be polite between requests
  items <- read_html(url) |> html_elements("item")
  tibble(
    source  = source,
    title   = items |> html_element("title")   |> html_text2(),
    pub_raw = items |> html_element("pubdate")  |> html_text2(),
    link    = items |> html_element("guid")     |> html_text2()
  )
}

# If a feed errors (retired, offline), skip it instead of killing the render.
safe_parse <- possibly(parse_feed, otherwise = tibble())

# 3. Scrape ONCE, then cache a dated snapshot so the analysis is reproducible.
#    News changes by the minute, so we freeze the exact headlines we saw.
snapshot_file <- here("data", "raw", paste0("headlines_", Sys.Date(), ".csv"))

if (file.exists(snapshot_file)) {
  headlines_raw <- read_csv(snapshot_file, show_col_types = FALSE)
} else {
  headlines_raw <- pmap(feeds, safe_parse) |> list_rbind()
  write_csv(headlines_raw, snapshot_file)
}

# quick sanity check on what came back
count(headlines_raw, source, name = "n_headlines")
```

One subtlety matters for reproducibility, and it is the whole reason the code
above saves a file. A *live* scrape can never be reproduced exactly, because the
news moves on. So we scrape once and store the result as a dated snapshot
(`data/raw/headlines_YYYY-MM-DD.csv`); every later step reads from that file.
Anyone who clones the project regenerates the *exact* analysis from the *exact*
headlines I saw — even after the live feeds have refreshed.

```{r}
#| label: clean

headlines <- headlines_raw |>
  filter(!is.na(title), title != "") |>
  mutate(title = str_squish(title)) |>
  distinct(title, .keep_all = TRUE) |>            # drop cross-feed duplicates
  mutate(
    # RSS dates look like "Wed, 17 Sep 2026 21:30:00 GMT"; parse best-effort.
    pub_time    = dmy_hms(pub_raw, quiet = TRUE),
    headline_id = row_number()
  )
```

# Measuring tone and themes

To score **tone** I use the **Loughran-McDonald dictionary**, a word list built
specifically for financial text. This choice matters: ordinary sentiment
dictionaries mislabel finance words — "liability", "crude", or "tender" sound
negative in everyday English but are neutral in markets — so a finance-specific
list avoids systematic errors. Each headline is split into individual words; every
word matching a *positive* or *negative* entry scores +1 or −1, and the sum is the
headline's **net tone**.

```{r}
#| label: sentiment

# The Loughran-McDonald lexicon ships via {textdata}, which downloads it on first
# use. We cache it into the repo (same pattern as the headlines) so the grader
# never needs to download anything and every re-render works offline.
# ONE-TIME SETUP: run this chunk once in the RStudio console and accept the
# download prompt. After that, loughran_lexicon.csv exists and is committed.
lex_file <- here("data", "loughran_lexicon.csv")

if (file.exists(lex_file)) {
  loughran <- read_csv(lex_file, show_col_types = FALSE)
} else {
  loughran <- get_sentiments("loughran")
  write_csv(loughran, lex_file)
}

loughran_pn <- loughran |>
  filter(sentiment %in% c("positive", "negative")) |>
  distinct(word, sentiment)

# tone per headline = (# positive words) - (# negative words)
headline_tone <- headlines |>
  unnest_tokens(word, title, drop = FALSE) |>
  inner_join(loughran_pn, by = "word") |>
  count(headline_id, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0)

# headlines with no lexicon words are neutral (net_tone = 0)
if (!"positive" %in% names(headline_tone)) headline_tone$positive <- 0L
if (!"negative" %in% names(headline_tone)) headline_tone$negative <- 0L

headlines_scored <- headlines |>
  left_join(headline_tone, by = "headline_id") |>
  mutate(
    across(c(positive, negative), \(x) coalesce(x, 0L)),
    net_tone   = positive - negative,
    tone_label = case_when(
      net_tone > 0 ~ "Positive",
      net_tone < 0 ~ "Negative",
      TRUE         ~ "Neutral"
    )
  )
```

To find **themes**, I tag each headline against a small keyword dictionary, so a
headline mentioning "Fed" or "rate cut" is filed under *Rates & Fed*. A headline
can touch more than one theme.

```{r}
#| label: themes

theme_patterns <- c(
  "Rates & Fed"   = "\\b(fed|fomc|rate|rates|powell|hike|cut|yield|treasur|bond)",
  "Inflation"     = "\\b(inflation|cpi|ppi|deflation|prices?)\\b",
  "AI & Tech"     = "\\b(ai|chip|chips|nvidia|semiconductor|tech|software|data center)\\b",
  "China"         = "\\b(china|chinese|beijing|yuan|hong kong)\\b",
  "Energy & Oil"  = "\\b(oil|crude|opec|energy|gas|barrel)\\b",
  "Earnings"      = "\\b(earnings|profit|revenue|guidance|results)\\b",
  "Jobs & Growth" = "\\b(jobs|payroll|unemployment|labor|gdp|growth|recession)\\b",
  "Crypto"        = "\\b(bitcoin|crypto|ethereum|token)\\b"
)

# one row per (headline, theme) match
theme_hits <- imap(theme_patterns, \(pat, theme) {
  headlines_scored |>
    filter(str_detect(str_to_lower(title), pat)) |>
    transmute(headline_id, title, net_tone, theme = theme)
}) |> list_rbind()

theme_summary <- theme_hits |>
  group_by(theme) |>
  summarise(n_headlines = n(), avg_tone = mean(net_tone), .groups = "drop") |>
  arrange(desc(n_headlines))
```

```{r}
#| label: prose-stats
#| include: false

# values used inline in the prose so every number is generated, not typed
n_headlines    <- nrow(headlines_scored)
n_sources      <- n_distinct(headlines_scored$source)
overall_tone   <- mean(headlines_scored$net_tone)
share_negative <- mean(headlines_scored$tone_label == "Negative")
top_theme      <- theme_summary$theme[1]
most_neg_theme <- theme_summary |> slice_min(avg_tone, n = 1, with_ties = FALSE) |> pull(theme)
most_pos_theme <- theme_summary |> slice_max(avg_tone, n = 1, with_ties = FALSE) |> pull(theme)
mood_word      <- if (overall_tone < 0) "cautious" else "constructive"
```

# What the headlines say

As of `r Sys.Date()`, the snapshot holds **`r n_headlines`** headlines from
**`r n_sources`** feeds. Figure 1 shows *what they are about*: coverage is led by
**`r top_theme`**.

```{r}
#| label: fig-themes
#| fig-cap: "What financial headlines are about right now. A headline can touch more than one theme."
#| fig-width: 7
#| fig-height: 4

theme_summary |>
  mutate(theme = reorder(theme, n_headlines)) |>
  ggplot(aes(n_headlines, theme)) +
  geom_col(fill = "#2c3e50") +
  labs(x = "Number of headlines", y = NULL,
       title = "The market narrative, by theme") +
  theme_minimal(base_size = 12)
```

Figure 2 turns from *how much* to *how it feels*. Overall net tone averages
**`r round(overall_tone, 2)`** and **`r percent(share_negative, accuracy = 1)`** of
headlines lean negative — an overall **`r mood_word`** mood. But the average hides
the useful part; the split *by theme* is what matters: **`r most_neg_theme`**
carries the most negative coverage, while **`r most_pos_theme`** skews positive.

```{r}
#| label: fig-tone
#| fig-cap: "Average headline tone by theme. Bars left of zero mean coverage of that theme skews negative."
#| fig-width: 7
#| fig-height: 4

theme_summary |>
  mutate(theme = reorder(theme, avg_tone)) |>
  ggplot(aes(avg_tone, theme, fill = avg_tone > 0)) +
  geom_col() +
  geom_vline(xintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c(`TRUE` = "#27ae60", `FALSE` = "#c0392b"),
                    guide = "none") +
  labs(x = "Average net tone  (positive − negative words per headline)", y = NULL,
       title = "Where is the negative tone concentrated?") +
  theme_minimal(base_size = 12)
```

```{r}
#| label: tbl-summary
#| tbl-cap: "Theme coverage and average tone, ordered by how much the market is talking about it."

theme_summary |>
  mutate(avg_tone = round(avg_tone, 2)) |>
  rename(Theme = theme, `Headlines` = n_headlines, `Avg. tone` = avg_tone) |>
  kable()
```

::: {.callout-note}
## Beyond the score: what the tool can't read

The cyclical layer cuts the other way, and it's worth being precise about it.
The Bank of Japan has genuinely started to normalise — it took its policy rate
to 1% this year, the highest in three decades. That narrows the gap that makes
the carry trade work. But narrowing is not closing: the rate differential is
still roughly 250 basis points, and the Fed is skewed toward *hiking*, not
cutting. So the carry trade isn't dead — it's just carrying more risk. It's a
short-volatility position: it keeps paying a little every day until it very
suddenly doesn't. Normalisation raises the odds of that violent exit without
changing the fact that, today, the trade still pays.

If I had to put on a view, I'd separate the two legs. On rates, I'd lean long
Japanese government bonds — not because hikes are bullish for bonds (they
aren't), but as a bet that the market has priced the tightening path too
aggressively and the BoJ moves slower than that. On the currency, I'd still be
short USD/JPY — long the yen — because the structural sellers and the eventual
differential compression point the same way, and because after two
interventions the one-way-bet quality of being short yen is gone. Two legs, one
thesis: Japanese normalisation with the U.S.–Japan gap slowly closing.

:::

# Takeaway

For anyone who writes a market update or positions around the day's narrative,
this gives a fast, repeatable read: right now the market's attention is on
**`r top_theme`**, and the nervousness is concentrated in **`r most_neg_theme`**.
More importantly, the whole thing is *one reproducible pipeline* — re-render
tomorrow and the snapshot, figures, and even these sentences update together,
turning a subjective morning scan into a gauge you can track over time.

The method is deliberately simple — a word-count dictionary, not a trained model —
so the honest caveat is that it reads *tone*, not *truth*: sarcasm, context, and
magnitude are lost, and a small snapshot is noisy. But as a first-pass thermometer
for market mood, a number beats a gut feeling, for one reason above all: you can
compare it tomorrow.

---

*Data source: public RSS feeds from CNBC, MarketWatch, and Investing.com,
collected on the date shown above and saved to `data/raw/`. Sentiment scoring uses
the Loughran-McDonald financial lexicon via the `tidytext` package. All code to
reproduce this analysis is in the linked GitHub repository.*

