---
title: "Did Paychecks Keep Up? Real Wages Through the 2021–2024 Inflation Shock"
date: 2026-09-28
categories:
  - labor market
  - inflation
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    df-print: kable
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(readr)      # read the CPS extract
library(dplyr)      # data wrangling
library(tidyr)      # reshaping
library(stringr)    # labels
library(forcats)    # ordered factors for plotting
library(ggplot2)    # figures
library(scales)     # axis formatting ($, %)
library(knitr)      # kable tables
library(here)       # project-relative paths

set.seed(6400)
options(scipen = 999)
```

# The question

Between 2021 and 2023 the United States lived through its sharpest burst of
inflation in forty years: consumer prices rose about 8% in 2022 alone. Wages
also rose over the same period, and headlines could point to "record pay
increases" on any given month. But a raise only makes you better off if it beats
the price of the things you buy. So the question that actually matters to a worker
is not *did my pay go up?* but **did my pay go up faster than prices?**

This post uses U.S. Census microdata to answer that for the American workforce:
**what happened to real (inflation-adjusted) hourly wages through the 2021–2024
inflation shock, and did every kind of worker fare the same?** The three figures
build the answer in layers: first the aggregate, then by education, then across
the wage distribution.

# The data and how it is measured

The data are individual-level records from the **IPUMS Current Population Survey
(CPS)**, Annual Social and Economic Supplement (ASEC), for survey years 2019 and
2021–2025. Each ASEC asks about income earned in the *previous* calendar year, so
these samples describe earnings for **2018 and 2020–2024**.

Two measurement choices matter and are worth stating plainly:

- **Hourly wage, not annual income.** Annual wage income mixes together how much
  someone earns *per hour* with how many hours they worked. To isolate the price
  of an hour of labour, I compute an hourly wage as
  `INCWAGE / (WKSWORK1 × UHRSWORKLY)` : annual wage income divided by weeks worked
  times usual hours per week.
- **Sampling weights.** The CPS is a *sample*, not a census, and each respondent
  stands in for a different number of real Americans. Every statistic below is
  therefore weighted by the ASEC person weight (`ASECWT`), so the numbers describe
  the U.S. working population, not just the people who happened to be surveyed.

To turn nominal dollars into **real** dollars I deflate every year's wage by the
BLS Consumer Price Index (CPI-U, annual average), expressing everything in **2024
dollars**. That way a wage in 2018 and a wage in 2024 can be compared on the same
purchasing-power ruler.

```{r}
#| label: load-and-build

# --- 1. Read the committed CPS extract (readr reads .gz directly) ----------
cps_raw <- read_csv(
  here("blog", "posts", "post3", "data", "cps_00001.csv.gz"),
  col_select = c(YEAR, AGE, SEX, EDUC, WKSWORK1, UHRSWORKLY, INCWAGE, ASECWT),
  show_col_types = FALSE
)

# --- 2. CPI-U annual average (BLS, series CUUR0000SA0), base = 2024 ---------
cpi <- tribble(
  ~income_year, ~cpi_u,
  2018, 251.107,
  2020, 258.811,
  2021, 270.970,
  2022, 292.655,
  2023, 304.702,
  2024, 313.689
)
cpi_base <- cpi$cpi_u[cpi$income_year == 2024]   # deflate everything to 2024 $

# --- 3. Build the analysis sample ------------------------------------------
# Keep prime-age (25-64) wage earners with valid, positive wages/weeks/hours.
# INCWAGE 99999998/99999999 = missing/NIU; UHRSWORKLY 999 = NIU — all dropped.
wages <- cps_raw |>
  mutate(income_year = YEAR - 1L) |>
  filter(
    AGE >= 25, AGE <= 64,
    INCWAGE > 0, INCWAGE < 99999998,
    WKSWORK1 >= 1, WKSWORK1 <= 52,
    UHRSWORKLY >= 1, UHRSWORKLY <= 99
  ) |>
  mutate(hourly_nom = INCWAGE / (WKSWORK1 * UHRSWORKLY)) |>
  filter(hourly_nom >= 2, hourly_nom <= 500) |>          # drop clear coding errors
  left_join(cpi, by = "income_year") |>
  mutate(
    hourly_real = hourly_nom * (cpi_base / cpi_u),         # 2024 dollars
    educ_group = case_when(
      EDUC %in% 2:72   ~ "Less than HS",
      EDUC == 73       ~ "High school",
      EDUC %in% 74:110 ~ "Some college",
      EDUC >= 111      ~ "Bachelor's+",
      TRUE             ~ NA_character_
    ),
    educ_group = factor(educ_group,
      levels = c("Less than HS", "High school", "Some college", "Bachelor's+"))
  )

# --- 4. Weighted quantile helper (self-contained, no extra package) ---------
wtd_quantile <- function(x, w, probs) {
  ok <- !is.na(x) & !is.na(w)
  x <- x[ok]; w <- w[ok]
  ord <- order(x); x <- x[ord]; w <- w[ord]
  cw <- (cumsum(w) - 0.5 * w) / sum(w)
  approx(cw, x, xout = probs, rule = 2)$y
}
wtd_median <- function(x, w) wtd_quantile(x, w, 0.5)
```

```{r}
#| label: aggregate

# Fig 1 data: weighted median wage per year, nominal vs real
agg_overall <- wages |>
  group_by(income_year) |>
  summarise(
    Nominal = wtd_median(hourly_nom,  ASECWT),
    Real    = wtd_median(hourly_real, ASECWT),
    .groups = "drop"
  ) |>
  pivot_longer(c(Nominal, Real), names_to = "measure", values_to = "wage")

# Fig 2 data: weighted median REAL wage per year, by education
agg_educ <- wages |>
  filter(!is.na(educ_group)) |>
  group_by(income_year, educ_group) |>
  summarise(real_median = wtd_median(hourly_real, ASECWT), .groups = "drop")

# Fig 3 data: weighted P10/P50/P90 of REAL wage, indexed to first year = 100
agg_pct <- wages |>
  group_by(income_year) |>
  summarise(
    P10 = wtd_quantile(hourly_real, ASECWT, 0.10),
    P50 = wtd_quantile(hourly_real, ASECWT, 0.50),
    P90 = wtd_quantile(hourly_real, ASECWT, 0.90),
    .groups = "drop"
  ) |>
  pivot_longer(c(P10, P50, P90), names_to = "pctile", values_to = "wage") |>
  group_by(pctile) |>
  arrange(income_year) |>
  mutate(index = 100 * wage / first(wage)) |>
  ungroup()
```

```{r}
#| label: prose-stats
#| include: false

n_workers   <- nrow(wages)
real_2018   <- agg_overall |> filter(income_year == 2018, measure == "Real") |> pull(wage)
real_2022   <- agg_overall |> filter(income_year == 2022, measure == "Real") |> pull(wage)
real_2024   <- agg_overall |> filter(income_year == 2024, measure == "Real") |> pull(wage)
nom_2018    <- agg_overall |> filter(income_year == 2018, measure == "Nominal") |> pull(wage)
nom_2024    <- agg_overall |> filter(income_year == 2024, measure == "Nominal") |> pull(wage)
nom_growth  <- 100 * (nom_2024 / nom_2018 - 1)
real_growth <- 100 * (real_2024 / real_2018 - 1)
real_dip    <- 100 * (real_2022 / real_2018 - 1)

# P10 vs P90 cumulative real change over the window
p10_end <- agg_pct |> filter(pctile == "P10") |> arrange(income_year) |> pull(index) |> last()
p90_end <- agg_pct |> filter(pctile == "P90") |> arrange(income_year) |> pull(index) |> last()
```

# What the data show

The analysis sample is **`r comma(n_workers)`** prime-age (25–64) wage earners
across the six years. Figure 1 sets up the puzzle.

```{r}
#| label: fig-nominal-real
#| fig-cap: "Median hourly wage, nominal vs. inflation-adjusted (2024 dollars). Nominal pay kept climbing; real pay stalled and fell during the 2021–22 inflation peak before recovering."
#| fig-width: 7.5
#| fig-height: 4.5

ggplot(agg_overall, aes(income_year, wage, colour = measure)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(Nominal = "#95a5a6", Real = "#2c3e50"), name = NULL) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_continuous(breaks = agg_overall$income_year) +
  labs(x = NULL, y = "Median hourly wage",
       title = "Nominal pay rose the whole time — real pay did not") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The two lines tell opposite stories. In nominal terms the median hourly wage rose
about **`r round(nom_growth)`%** from 2018 to 2024: the "pay is rising" headline.
But once we strip out inflation, the real median wage in 2022 was actually
**`r sprintf('%+.1f', real_dip)`%** relative to 2018: the raises workers received
were smaller than the price increases they faced, so purchasing power *fell* at the
height of the inflation shock. By 2024 real wages had clawed back to about
**`r sprintf('%+.1f', real_growth)`%** versus 2018: a recovery, but a modest one
spread over six years.

Figure 2 asks *who* absorbed the squeeze, by splitting the real median wage by
education.

```{r}
#| label: fig-education
#| fig-cap: "Median real hourly wage (2024 dollars) by education. The gap between more- and less-educated workers is wide and persistent, and every group dips during the inflation peak."
#| fig-width: 7.5
#| fig-height: 4.5

ggplot(agg_educ, aes(income_year, real_median, colour = educ_group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_colour_brewer(palette = "Set2", name = NULL) +
  scale_y_continuous(labels = label_dollar()) +
  scale_x_continuous(breaks = unique(agg_educ$income_year)) +
  labs(x = NULL, y = "Median real hourly wage (2024 $)",
       title = "The inflation dip hit every education group") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The levels are unsurprising, more schooling, higher pay. But the *shape* is the
point: all four lines bend downward through 2021–2022 together. Inflation did not
respect a diploma; it lowered real pay across the board before the partial
recovery. That pushes the interesting question one level deeper: within the
workforce, did the *lowest*-paid workers lose the most? Figure 3 answers it by
tracking the 10th, 50th, and 90th percentiles of the real wage, each indexed to
its own 2018 value.

```{r}
#| label: fig-percentiles
#| fig-cap: "Cumulative change in real hourly wage by percentile, indexed to 2018 = 100. A line below 100 means that part of the distribution had less purchasing power than in 2018."
#| fig-width: 7.5
#| fig-height: 4.5

ggplot(agg_pct, aes(income_year, index, colour = pctile)) +
  geom_hline(yintercept = 100, colour = "grey60", linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(P10 = "#c0392b", P50 = "#2c3e50", P90 = "#27ae60"),
                      name = NULL) +
  scale_x_continuous(breaks = unique(agg_pct$income_year)) +
  labs(x = NULL, y = "Real wage (2018 = 100)",
       title = "Who ended up ahead of inflation, and who didn't") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

```{r}
#| label: tbl-pct
#| tbl-cap: "Real hourly wage percentiles (2024 dollars), by income year."

agg_pct |>
  mutate(wage = round(wage, 2)) |>
  select(income_year, pctile, wage) |>
  pivot_wider(names_from = pctile, values_from = wage) |>
  rename(`Income year` = income_year) |>
  kable()
```

By the end of the window the bottom of the distribution (P10) sits near
**`r round(p10_end)`** on the 2018 = 100 scale, while the top (P90) sits near
**`r round(p90_end)`**. Read together, the three figures say the inflation shock
was broadly shared rather than sharply regressive: every group's real wage dipped
and then largely recovered, and low-wage workers did *not* fall dramatically
further behind high-wage workers over the full period.

# Takeaway

The headline "wages are rising" was true and misleading at the same time. Nominal
pay rose throughout, but the CPS microdata show that **real** median pay fell
during 2021–2022 (workers' raises lost a race against prices) and only recovered
to modestly above its 2018 level by 2024. Splitting the workforce shows the squeeze
was general: it hit every education group and ran across the wage distribution,
rather than concentrating on one part of it.

The broader lesson is methodological, and it is the whole reason this post uses
weighted microdata rather than a single headline number: to say something true
about *the American worker*, you have to (1) adjust nominal pay for inflation and
(2) weight the sample so it represents the population. Do neither and "pay is up
8%" reads as good news. Do both and you see that, for the typical worker, the
inflation years were a real-wage setback that took until 2024 to undo.

---

*Data source: IPUMS CPS (Flood et al.), ASEC samples for survey years 2019 and
2021–2025, extract in `data/cps_00001.csv.gz`. Inflation adjustment uses the BLS
Consumer Price Index for All Urban Consumers (CPI-U, series CUUR0000SA0, annual
average). All statistics use the ASEC person weight (ASECWT). Code to reproduce
the analysis is in the linked GitHub repository.*
